---
title: "ACRV"
output: html_document
date: "2024-03-14"
---

```{r setup, include=FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(readr)
library(reshape2)
library(ape)
library(Claddis)
library(vioplot)
library(phangorn)
```

First, let's pull some basics - what model was preferred by what dataset?
```{r include=FALSE}
model_indicator <- c()
alpha_values <-c()
alpha_indicators <-c()
names <-c()
median_pvalues <-c()
median_lvalues <- c()
sr1 <- c()
sr2 <- c()
sr3 <- c()
sr4 <- c()
diffs <-c()
setwd("2016_and_below")
for (file in list.files(path=".", pattern=".log")){
  dataf <- read_delim(file, show_col_types = FALSE)
  rates_ind = sum(dataf$site_rates_indicator)
  model_indicator = c(model_indicator, rates_ind)
  names <- c(names, file)
  median = median(dataf$alpha)
  med_ind = median(dataf$alpha_indicator)
  alpha_values <- c(alpha_values, median)
  alpha_indicators <- c(alpha_indicators, med_ind)
  medianP = median(dataf$Posterior)
  median_pvalues <- c(median_pvalues, medianP)
  medianL = median(dataf$Likelihood)
  median_lvalues <- c(median_lvalues, medianL)
  median1 = median(dataf$`site_rates[1]`)
  median2 = median(dataf$`site_rates[2]`)
  median3 = median(dataf$`site_rates[3]`)
  median4 = median(dataf$`site_rates[4]`)
  median_vec <- c(median1, median2, median3, median4)
  min_med <- min(median_vec)
  max_med <- max(median_vec)
  diffs <- c(diffs, max_med - min_med)
  mv1 <- sort(median_vec)[1]
  mv2 <- sort(median_vec)[2]
  mv3 <- sort(median_vec)[3]
  mv4 <- sort(median_vec)[4]
  sr1 <- c(sr1, mv1)
  sr2 <- c(sr2, mv2)
  sr3 <- c(sr3, mv3)
  sr4 <- c(sr4, mv4)
}
```

Wrap it all up in a dataframe:
```{r}

alpha_indicators <- alpha_indicators[2:570]
names <- names[2:570]
alpha_values <- alpha_values[2:570]
diff_values <- diffs[2:570]
median_pvalues <-median_pvalues[2:570]
median_lvalues <- median_lvalues[2:570]
sr1 <- sr1[2:570]
sr2 <- sr2[2:570]
sr3 <- sr3[2:570]
sr4 <- sr4[2:570]

dataframe <- data.frame(names, model_indicators, alpha_indicators, alpha_values, median_pvalues, median_lvalues, diff_values, sr1, sr2, sr3, sr4)

```

First, model preferences. Let's start with preferring one model over the other:

```{r}

ggplot(data = dataframe, aes(x = model_indicator)) + geom_histogram() + xlab("Preference Between Gamma and Lognormal") + geom_vline(xintercept = 1.5) + theme_bw() +
  annotate("text",               
          x = 1.25, y = 225,         
           label = "Prefer Gamma Model") +
  annotate("text",               
          x = 1.8, y = 225,         
           label = "Prefer Lognormal Model")


```
Well, shoot. A ton of these matrices prefer the Lognormal Model. Let's pull those out so we can test the effects of model misspecification if we use the Gamma model on them.

```{r}

library(phangorn)

log_pref <-dataframe %>% 
  filter(model_indicator > 1.5)

files <- log_pref$names
#Remove some files that had malformed outputs 
files <- files[files != 'Bai.log']
files <- files[files != "de_Pinna_etal_2007a.nex.clear.log"]
files <- files[files != "Gasulla_etal_2015aa.nex.clear.log"]
files <- files[files != "Ksepka_et_Clarke_2010a.nex.clear.log"]
files <- files[files != "Langer_etal_2014a.nex.clear.log"]
files <- files[files != "Ni_etal_2016ba.nex.clear.log"]
files <- files[files != "Nicholls_1999a.nex.clear.log"]
files <- files[files != "Norell_1989a.nex.clear.log"]
files <- files[files != "Ruta_et_Coates_2003a.nex.clear.log"]
files <- files[files != "Shirai_1996a.nex.clear.log"]
files <- files[files != "Spielman_etal_2008aa.nex.clear.log"]
files <- files[files != "Tanaka_etal_2018b.nex.clear.log"]
files <- files[files != "Tattersall_1993a.nex.clear.log"]
files <- files[files != "Tsuji_2017a.nex.clear.log"]
files <- files[files != "Tsuji_etal_2013b.nex.clear.log"]
files <- files[files != "Tykoski_etal_2019a.nex.clear.log"]
files <- files[files != "Valin_et_Laurin_2004a.nex.clear.log"]
files <- files[files != "Vari_1995a.nex.clear.log"]
files <- files[files != "Verzi_etal_inpressa.nex.clear.log"]
files <- files[files != "Verzi_etal_inpressa.nex.clear.log"]
files <- files[files != "Vigliotta_2008a.nex.clear.log"]
files <- files[files != "Vincent_et_Storrs_2019ab.nex.clear.log"]
files <- files[files != "Vlachos_et_Rabi_inpressa.nex.clear.log"]
files <- files[files !=  "Zanata_et_Vari_2005a.nex.clear.log"]
files <- files[files !=  "Zalmout_etal_2010a.nex.clear.log"]
files <- files[files !=  "Zanno_et_Makovicky_2011ab.nex.clear.log"]
files <- files[files !=  "Zhu_et_Schultze_2001a.nex.clear.log"]

dists <- c()
b_score <- c()
diffs <- c()


for (file in files){
  filename <-  unlist(strsplit(file, "[.]"))
  gname <- paste0("logs_gamma_only/", filename[1], ".", filename[2], ".final.mcc.tre")
  lname <- paste0("2016_and_below/", filename[1], ".", filename[2], ".", filename[3], ".trees.mcc.tre")

  gamma_tree <- read.nexus(gname)
  lognorm_tree <- read.nexus(lname)
  dists <- c(RF.dist(gamma_tree, lognorm_tree, normalize = TRUE), dists)
  b_score <- c(wRF.dist(gamma_tree, lognorm_tree), b_score)
  diffs <- c(sum(gamma_tree$edge.length) - sum(lognorm_tree$edge.length), diffs)
}

rf <- data.frame(dists, b_score, diffs)  


ggplot(data = rf, aes(x = dists)) + geom_histogram() + theme_bw() + xlim(0, 1) + ylim(0, 30) +
   labs(title = "Misspecified Model: RF Distance",
         x = "Scaled RF distance between Gamma and Lognormal trees",
         y = "Count") +
  annotate("text",               
          x = 0.2, y = 30,         
           label = "Total Datasets = 318") 
ggsave("GammaLogRF.png")

ggplot(data = rf, aes(x = diffs)) + geom_histogram() + theme_bw() + xlim(-10, 10) + ylim(0, 30) +
   labs(title = "Misspecified Model: Branch Length Difference",
         x = "Branch Length distance between Gamma and Lognormal trees",
         y = "Count") +
  annotate("text",               
          x = -5, y = 30,         
           label = "Lognormal Trees Longer") +
  annotate("text",               
          x = 5, y = 30,         
           label = "Gamma Trees Longer") +
   geom_vline(xintercept = 0)
ggsave("GammaLogBscore.png")

```



```{r}
#dataframes that are filtered into ACRV being favored (alpha indicator=1) or not (alpha indicator=0)
acrv_0 <- dataframe %>%
  filter(alpha_indicators == 0)
acrv_0.5 <- dataframe %>%
  filter(alpha_indicators == 0.5)
acrv_1 <- dataframe %>%
  filter(alpha_indicators == 1)

#model preference
model_prefs <- dataframe %>% 
  count(alpha_indicators == 1)

filter_100 <- dataframe %>% 
  filter(alpha_values > 100)

filtered_alpha0 <- dataframe %>%
  filter(alpha_indicators != 1)

filtered_diffs <- dataframe %>% 
  filter(alpha_indicators != 1.0) 

filtered_acrv <- dataframe %>% 
  filter(alpha_indicators == 1) 

filtered_g4 <- dataframe %>%
  filter(alpha_indicators == 1)

d3 <- density(filtered_g4$diff_values)
deng1_1 <- density(as.vector(filtered_g4$sr1))
deng2_1 <- density(as.vector(filtered_g4$sr2))
deng3_1 <- density(as.vector(filtered_g4$sr3))
deng4_1 <- density(as.vector(filtered_g4$sr4))

sr_df <- melt(dataframe[,5:8])
g1 <- as.vector(dataframe$sr1)
g2 <- as.vector(dataframe$sr2)
g3 <- as.vector(dataframe$sr3)
g4 <- as.vector(dataframe$sr4)
deng1 <- density(as.vector(dataframe$sr1))
deng2 <- density(as.vector(dataframe$sr2))
deng3 <- density(as.vector(dataframe$sr3))
deng4 <- density(as.vector(dataframe$sr4))

# Quick count - 385 of 577 datasets favor Rate Het


```

Now, for comparisons. First, let's look at the effects of not including ACRV. If the dataset favors ACRV, but it is not included, we find it in the no_acrv_acrv_favored directory.

```{r include=FALSE}

alpha_values <-c(0)
alpha_indicators <-c(0)
names <-c(0)
median_pvalues <-c(0)
median_lvalues <- c(0)
setwd("no_acrv_acrv_favored/")
for (file in list.files(path=".", pattern=".log")){
  dataf <- read_delim(file, show_col_types = FALSE)
  names <- c(names, file)
    medianP = median(dataf$Posterior)
    median_pvalues <- c(median_pvalues, medianP)
    medianL = median(dataf$Likelihood)
    median_lvalues <- c(median_lvalues, medianL)
}

names <- names[2:401]
median_pvalues <- median_pvalues[2:401]
median_lvalues <- median_lvalues[2:401]

no_acrv_dataframe_p <- data.frame(names,median_pvalues)
no_acrv_dataframe_l <- data.frame(names,median_lvalues)

```

We can merge this to the dataframe of posterior probabilities for the ACRV sets. Then, we can calculate how badly it messes up the PP to not have ACRV:

```{r}
 pvalue_dataframe <- dataframe %>% 
  select(names, median_pvalues)

 lvalue_dataframe <- dataframe %>%
   select(names, median_lvalues)

 no_acrv_dataframe_p <- no_acrv_dataframe_p %>% 
  select(names, median_pvalues) %>% 
   rename("no_acrv_pvalues" = median_pvalues)
 
 no_acrv_dataframe_l <- no_acrv_dataframe_l %>%
   select(names, median_lvalues) %>%
    rename("no_acrv_lvalues" = median_lvalues)

joined_p <- merge(pvalue_dataframe, no_acrv_dataframe_p, by = "names", )
joined_l <- merge(lvalue_dataframe, no_acrv_dataframe_l, by = "names", )

new_p <- joined_p %>%
  mutate(diff = median_pvalues - no_acrv_pvalues)
new_l <- joined_l %>%
  mutate(diff = median_lvalues - no_acrv_lvalues)
```
And now the ones where ACRV was un favored and left off:
```{r include=FALSE}

alpha_values <-c(0)
alpha_indicators <-c(0)
names <-c(0)
median_pvalues <-c(0)
median_lvalues <- c(0)
setwd("no_het-preferNoHet/log/")
for (file in list.files(path=".", pattern=".log")){
  dataf <- read_delim(file, show_col_types = FALSE)
  names <- c(names, file)
    medianP = median(dataf$Posterior)
    median_pvalues <- c(median_pvalues, medianP)
    medianL = median(dataf$Likelihood)
    median_lvalues <- c(median_lvalues, medianL)
}

names <- names[2:153]
median_pvalues <- median_pvalues[2:153]
median_lvalues <- median_lvalues[2:153]

unfavoreddataframe_p <- data.frame(names,median_pvalues)
unfavoreddataframe_l <- data.frame(names,median_lvalues)

unfavoreddataframe_p <- unfavoreddataframe_p %>% 
  select(names, median_pvalues) %>% 
   rename("no_acrv_pvalues" = median_pvalues)

unfavoreddataframe_l <- unfavoreddataframe_l %>% 
  select(names, median_lvalues) %>% 
   rename("no_acrv_lvalues" = median_lvalues)

joined2_p <- merge(pvalue_dataframe, unfavoreddataframe_p, by = "names", )
joined2_l <- merge(lvalue_dataframe, unfavoreddataframe_l, by = "names", )

new2_p <- joined2_p %>%
  mutate(diff = median_pvalues - no_acrv_pvalues)
new2_l <- joined2_l %>%
  mutate(diff = median_lvalues - no_acrv_lvalues)
```


```{r}
trees_no_acrv <- list.files(path="no_acrv_acrv_favored/", pattern="mcc.tre")
trees_acrv <- list.files(path="rjMCMC_new/", pattern="mcc.tre")
shared_trees <- trees_no_acrv[trees_no_acrv %in% trees_acrv]
shared_trees <- shared_trees[shared_trees != "Boyer_etal_2018a.nex.final.mcc.tre"]
shared_trees <- shared_trees[shared_trees != "Ely_et_Case_inpressa.nex.final.mcc.tre"]
shared_trees <- shared_trees[shared_trees != "Evans_etal_2012a.nex.final.mcc.tre"]
shared_trees <- shared_trees[shared_trees != "Femenias-Gual_etal_2017a.nex.final.mcc.tre"]
shared_trees <- shared_trees[shared_trees != "Fiorillo_et_Tykoski_2014ab.nex.final.mcc.tre"]
shared_trees <- shared_trees[shared_trees != "Foffa_etal_inpressa.nex.final.mcc.tre"]
shared_trees <- shared_trees[shared_trees != "Geroto_et_Bertini_inpressa.nex.final.mcc.tre"]
shared_trees <- shared_trees[shared_trees != "Groves_et_Eaglen_1988a.nex.final.mcc.tre"]
shared_trees <- shared_trees[shared_trees != "Halliday_etal_2016a.nex.final.mcc.tre"]
shared_trees <- shared_trees[shared_trees != "Halliday_etal_inpressba.nex.final.mcc.tre"]
shared_trees <- shared_trees[shared_trees != "Halliday_etal_inpressbb.nex.final.mcc.tre"]
shared_trees <- shared_trees[shared_trees != "Longrich_etal_2012a.nex.final.mcc.tre"]

dists <- c(0)
b_score <- c(0)
diffs <- c(0)

for (tree in shared_trees){
  tree
    acrv_path <- paste0("rjMCMC_new/", tree)
    no_acrv_path <- paste0("no_acrv_acrv_favored/", tree)
    acrv_tree <- read.nexus(acrv_path)
    no_acrv_tree <- read.nexus(no_acrv_path)
    dists <- c(RF.dist(acrv_tree, no_acrv_tree, normalize = TRUE), dists)
    b_score <- c(wRF.dist(acrv_tree, no_acrv_tree), b_score)
    diffs <- c(sum(acrv_tree$edge.length) - sum(no_acrv_tree$edge.length), diffs)
}

rf <- data.frame(dists, b_score, diffs)
```


Figures 
Figure 1
```{r}
par(mfrow=c(2,1),
    mar=c(4,4,1.5,0.75))
hist(dataframe$alpha_indicators, main="", xlim=c(0,1), ylim=c(0,450), xlab="ACRV Preference", ylab = "Count", yaxs="i")
mtext("A", adj=-0.125, line=0, cex=2)
mtext("Strong Preference:
      No ACRV", adj=0.04, line=-3.55, cex=0.55)
mtext("Strong Preference:
    ACRV            ", adj=0.975, line=-0.95, cex=0.55)

hist(dataframe$alpha_values, main="", yaxs="i", ylab = "Count", xlab = "Alpha Value",
     xlim=c(0, 1e+08), ylim=c(0, 300), breaks = 50)
mtext("B", adj=-0.125, line=0, cex=2)
text(0.20e+07, 195, "Strong ACRV", cex=0.55)
text(9.85e+07, 280, "No ACRV", cex=0.55)
```

Figure 2
```{r}
nf <- layout( matrix(c(1,3,2,4), nrow=2, byrow=TRUE) )
par(mar=c(4, 4, 2, 2))

#A
hist(0, yaxs="i", xlim=c(0,2.5), ylim=c(0,250), lwd=1,
     main = "ACRV Favored", xlab="Gamma Rate", ylab="Count")
hist(filtered_acrv$sr1, lwd=1, col= adjustcolor("#B78686", alpha.f = 0.4), breaks = 15, add=TRUE)
hist(filtered_acrv$sr2, lwd=1, col= adjustcolor("#E6E3A9", alpha.f = 0.4), add=TRUE)
hist(filtered_acrv$sr3, lwd=1, col= adjustcolor("#A699BF", alpha.f = 0.4), breaks=2, add=TRUE)
hist(filtered_acrv$sr4, lwd=1, col= adjustcolor("#85B7B4", alpha.f = 0.4), breaks=20, add=TRUE)
mtext("A", adj=-0.25, line=0, cex=2)
#C
hist(filtered_acrv$diff_values, xlab="Maximum Gamma - Minimum Gamma", 
     ylab="Count", main="", yaxs="i", ylim=c(0,250), col=adjustcolor("black", alpha.f = 0.2))
mtext("C", adj=-0.25, line=0, cex=2)
#B
hist(c(filtered_alpha0$sr1, filtered_alpha0$sr2),
     main = "ACRV Unfavored", xlab = "Gamma Rate", ylab="",
     yaxs="i", xlim=c(0.9998, 1.0002), ylim=c(0,200), col = c(adjustcolor("#B78686", alpha.f = 0.4), adjustcolor("#E6E3A9", alpha.f = 0.4)))
hist(c(filtered_alpha0$sr3, filtered_alpha0$sr4), col= c(adjustcolor("#A699BF", alpha.f = 0.4), adjustcolor("#85B7B4", alpha.f = 0.4)), add = TRUE)
mtext("B", adj=-0.25, line=0, cex=2)
#D
hist(filtered_alpha0$diff_values, xlim=c(0,0.005), main="", yaxs="i", 
     xlab="Maximum Gamma - Minimum Gamma", ylab="", col=adjustcolor("black", alpha.f = 0.2))
mtext("D", adj=-0.25, line=0, cex=2)
```

Figure 3 (posterior probabilities)
```{r}
par(mfrow=c(2,1),
    mar=c(4,4,1.5,0.75))

hist(0, main="ACRV favored but not modeled", yaxs="i", xaxs="i", ylab = "Count", xlab = "Posterior probability",
     xlim=c(-50, 1000), ylim=c(0, 300))
hist(new_p$diff, lwd=1, breaks=2500, add=TRUE)
mtext("A", adj=-0.125, line=0, cex=2)

hist(0, main="ACRV unfavored and not modeled", yaxs="i", xaxs="i", ylab = "Count", xlab = "Posterior probability",
     xlim=c(-9250, 2250), ylim=c(0, 25))
hist(new2_p$diff, lwd=1, breaks=100, add=TRUE)
mtext("B", adj=-0.125, line=0, cex=2)
```

Figure 3 (Likelihood)
```{r}
par(mfrow=c(2,1),
    mar=c(4,4,1.5,0.75))

hist(0, main="ACRV favored but not modeled", yaxs="i", xaxs="i", ylab = "Count", xlab = "Likelihood",
     xlim=c(-50, 1000), ylim=c(0, 175))
hist(new_l$diff, lwd=1, breaks=2500, add=TRUE)
mtext("A", adj=-0.125, line=0, cex=2)

hist(0, main="ACRV unfavored and not modeled", yaxs="i", xaxs="i", ylab = "Count", xlab = "Likelihood",
     xlim=c(-9250, 2250), ylim=c(0, 20))
hist(new2_l$diff, lwd=1, breaks=100, add=TRUE)
mtext("B", adj=-0.125, line=0, cex=2)
```
```{r}
par(mfrow=c(3,1),
    mar=c(4,4,1.5,0.75))

hist(rf$dists, breaks=100, main="Robinson-Foulds Metrics", ylab="Count", xlab="RF Distance", yaxs="i", ylim=c(0,125))
mtext("A", adj=-0.075, line=-0.75, cex=2)
hist(rf$b_score, breaks=100, main="", ylab="Count", xlab="RF B-Score", yaxs="i", ylim=c(0,175))
mtext("B", adj=-0.075, line=0, cex=2)
hist(rf$diffs, breaks=100, main="", ylab="Count", xlab="RF Differences", yaxs="i", ylim=c(0,175))
mtext("C", adj=-0.075, line=0, cex=2)
```
